---
title: "Lab 2: Distances and Projections"
subtitle: 'ESS 523c: Environmental Data Science Applications'
author:
  - name: Alexa Luna-Turner
    email: alexa12@colostate.edu
format: html
---

Libraries

```{r}
pacman::p_load(
  tidyverse, sf, units, tigris, rnaturalearth, gghighlight, ggrepel, flextable
)

options(tigris_use_cache = TRUE)
```

## Question 1: Build Spatial Dataset

1.1 Define a Projection

```{r}
eqdc <- '+proj=eqdc +lat_0=40 +lon_0=-96 +lat_1=20 +lat_2=60 +x_0=0 +y_0=0 +datum=NAD83 +units=m +no_defs'
```

1.2 CONUS State Boundary

```{r}
states_us <- states(cb = TRUE, progress_bar = FALSE) |>
  filter(!STUSPS %in% c("AK", "HI", "PR", "GU", "VI", "MP", "AS", "DC"))

conus_boundary <- st_transform(states_us, eqdc)
#Data from year 2021
```

1.3 North American Country Boundary

```{r}
na_countries <- ne_countries(returnclass = "sf") |>
  filter(admin %in% c("United States of America", "Canada", "Mexico"))

na_country_boundaries <- st_transform(na_countries, eqdc)
```

1.4 US City Locations

```{r}
cities <- readr::read_csv("data/uscities.csv") |>
  st_as_sf(coords = c("lng", "lat"), crs = 4326) #Convert to sf object

cities <- st_transform(cities, eqdc) #match projection

us_cities <- cities |>
  st_filter(st_union(conus_boundary))
```

Verify

```{r}
# All three must return the same CRS
st_crs(conus_boundary) == st_crs(na_country_boundaries)
st_crs(conus_boundary) == st_crs(us_cities)

# Row counts should be reasonable
nrow(conus_boundary)    # should be 48
nrow(us_cities)    # should be ~28,000

#Note for future reference
#lab labels - my labels
###states - conus_boundary
###contries - na_country_boundaries
### cities - us_cities
```

## Question 2: Distance Calculations

2.1 Distance to US National Boarder (km)

```{r}
resolved_boundary <- st_union(conus_boundary) |>
  st_boundary() |>
  st_cast("MULTILINESTRING")

us_cities <- us_cities |>
  mutate(dist_to_border = st_distance(geometry, resolved_boundary) |>
           set_units("km") |>
           as.numeric())

us_cities |>
  st_drop_geometry() |>
  arrange(desc(dist_to_border)) |>
  slice_head(n = 5) |>
  select(city_ascii, state_name, dist_to_border) |>
  flextable() |>
  set_header_labels(
    city_ascii = "City",
    state_name = "State",
    dist_to_border = "Distance to Border (km)")
```

2.2 Distance to Nearest State Boarder (km)

```{r}
preserved_boundary <- st_combine(conus_boundary) |>
  st_boundary() |>
  st_cast("MULTILINESTRING")

us_cities <- us_cities |>
  mutate(dist_to_state = st_distance(geometry, preserved_boundary) |>
           set_units("km") |>
           as.numeric())

us_cities |>
  st_drop_geometry() |>
  arrange(desc(dist_to_state)) |>
  slice_head(n = 5) |>
  select(city_ascii, state_name, dist_to_state) |>
  flextable() |>
  set_header_labels(
    city_ascii = "City",
    state_name = "State",
    dist_to_state = "Distance to State Border (km)")
```

2.3 Distance to Mexico (km)

```{r}
#function for next questions
country_distance <- function(country_name, new_col_name) {

  country_border <- na_country_boundaries |>
    filter(admin == country_name) |>
    st_boundary() |>
    st_cast("MULTILINESTRING")

  dist <- st_distance(us_cities$geometry, country_border) |>
    units::set_units("km") |>
    as.numeric()

  us_cities <<- us_cities |>
    mutate(!!new_col_name := dist)
  }

country_distance("Mexico", "dist_to_mexico")

us_cities |>
  st_drop_geometry() |>
  arrange(desc(dist_to_mexico)) |>
  slice_head(n = 5) |>
  select(city_ascii, state_name, dist_to_mexico) |>
  flextable() |>
  set_header_labels(
    city_ascii = "City",
    state_name = "State",
    dist_to_mexico = "Distance to Mexico (km)")
```

2.4 Distance to Canada (km)

```{r}
country_distance("Canada", "dist_to_canada")

us_cities |>
  st_drop_geometry() |>
  arrange(desc(dist_to_canada)) |>
  slice_head(n = 5) |>
  select(city_ascii, state_name, dist_to_canada) |>
  flextable() |>
  set_header_labels(
    city_ascii = "City",
    state_name = "State",
    dist_to_canada = "Distance to Canada (km)")
```

## Question 3: Mapping Distance Patterns

3.1 Reference Map

```{r}
ggplot() +
  geom_sf(data = na_country_boundaries, fill = "grey80", color = NA) + #North Amer
  geom_sf(data = conus_boundary, fill = NA, color = "black", linetype = "dashed", linewidth = 0.3) + #State boudaries
  geom_sf( data = us_cities |>
  slice_max(population, n = 10), color = "red", size = 2) + #Largest cities
  ggrepel::geom_label_repel(data = us_cities |>
      slice_max(population, n = 10),
      aes(geometry = geometry, label = city),
    stat = "sf_coordinates", size = 3)
```

3.2 Distance to National Boarder

```{r}
plot_distance_map <- function(distance_col, legend_name) {
 
   ggplot() +
    geom_sf(data = conus_boundary, fill = NA, color = "black", linewidth = 0.4) + #CONUS boudaries
    
    geom_sf(data = us_cities, #colored cities
      aes(color = .data[[distance_col]]), size = 1) +
    
    geom_sf(
      data = us_cities |> #top 5 furthest
      slice_max(order_by = .data[[distance_col]], n = 5), color = "red", size = 2) +
    
    ggrepel::geom_label_repel(
      data = us_cities |>
        slice_max(order_by = .data[[distance_col]], n = 5),
      aes(geometry = geometry, label = city),
      stat = "sf_coordinates",
      size = 3) +
      scale_color_viridis_c(
      name = legend_name)
}

plot_distance_map("dist_to_border", "Distance to National Border")
```

In 2–3 sentences: (1) where are the cities farthest from the national border? (2) Does the geographic pattern make sense given the shape of CONUS? (3) Name one water resources application where distance to a national border would be a meaningful variable.

According to the first flextable and the map above, all 5 cities that are furthest from the nation border are in Kansas. Yes the geographic pattern makes sense as the blue follows the borders of the country. If a water resource crosses national borders (which it often does), then certain policies and restrictions may be in place, so knowing the distance to the border can help policy making and infrastructure plans.

3.3 Distance to Nearest State Border

```{r}
plot_distance_map("dist_to_state", "Distance to State Border")
```

In 2–3 sentences: (1) how does the pattern differ from 3.2? (2) Which states appear to have the largest interior areas — do any water resources features (large river basins, aquifer extents) correlate with the states where cities can be far from borders? (3) What would a hydrologist use this distance for?

In this map, all 5 cities are in Texas instead of Kansas, also the distance pattern is different, following the state borders instead of the national border. Larger or more uniform (boxed) states have the largest interiors and there are several large aquifer in Texas where the 5 cities are located. Many water resources also cross state boundaries, so knowing the distance to that border for every city can help idenify cities that rely on/affect water resources across multiple jurisdictions.

3.4 Equidistant Zone: Mexico and Canada

```{r}
us_cities <- us_cities |>
  mutate(difference_mexico_canada = abs(dist_to_mexico - dist_to_canada))

ggplot() +
  geom_sf(data = conus_boundary, fill = NA,  color = "black",  linewidth = 0.4) +
  
  geom_sf(data = us_cities,
    aes(color = difference_mexico_canada), size = 1) +
  
  scale_color_viridis_c() +
  gghighlight(difference_mexico_canada < 100) +
  
  ggrepel::geom_label_repel(
  data = us_cities |>
    filter(difference_mexico_canada < 100) |>
    slice_max(population, n = 7),
  aes(geometry = geometry, label = city_ascii),
  stat = "sf_coordinates",
  size = 3)
```

In 2–3 sentences: (1) describe where the equidistant zone runs geographically, (2) does it surprise you that this band cuts through the interior of the country where it does? (3) What does this zone imply about the relative influence of Canadian vs. Mexican climate systems on cities in this corridor?

The equidistant zone runs approximately in the middle of the US, moving up or down as the national borders shift which does not surprise me. Canada and Mexico respectively will have more influence where this corridor is closer the their borders.

## Question 4: Real World Application- The 100 Mile Border Zone

4.1 Quantify the Zone

```{r}
summary_stats <- us_cities |>
  st_drop_geometry() |>
  summarize(
    city_count = sum(dist_to_border <= 160, na.rm = TRUE),
    population_in_zone = sum(population[dist_to_border <= 160], na.rm = TRUE),
    total_population = sum(population, na.rm = TRUE)) |>
  mutate(
    percent_of_total = (population_in_zone / total_population) * 100) |>
  select(-total_population)

summary_stats |>
  flextable() |>
  set_header_labels(
    city_count = "Number of Cities (Within 160 km)",
    population_in_zone = "Population in Border Zone",
    percent_of_total = "Percent of Total Population (%)")

#Does your result match the ACLU’s published estimate?
##ACLU -> 200 million people
## My estimate -> 258 million
###Yes, this rough estimate is close to the ACLU's.
```

In 2–3 sentences: (1) report the percentage and compare it to the ACLU’s estimate, (2) explain why the result might be higher than many people expect given the geography of the US, (3) name one scenario where a water resources professional might need to navigate federal jurisdictional boundaries in a similar way — where a regulatory zone defined by distance from a physical boundary changes what analysis or permitting is required.

According the the table, about 65% of the population lives in the 100 mile zone which is close the ACLU's statement that around two-thirds (66%) of the population is within the zone. Many people probably expect most of the population to be spread across the US given its size instead of clustered on the coast. Water resources crossing boundaries and near communities may require more aalysis or permits for safety reasons given the proximity to cities.

4.2 Map the Zone

```{r}
us_cities <- us_cities |>
  mutate(dist_to_border_num = as.numeric(dist_to_border))

zone_cities <- us_cities |>
  filter(dist_to_border_num < 160)

ggplot() +
  geom_sf(data = conus_boundary, fill = NA, color = "black", linewidth = 0.4) +
  
  geom_sf(data = us_cities |>
    filter(dist_to_border_num < 160),
    aes(color = dist_to_border_num), size = 1) +
  
  scale_color_gradient(low = "orange", high = "darkred") +
  
  gghighlight(dist_to_border_num < 160) +
  
  ggrepel::geom_label_repel(
    data = us_cities |>
      filter(dist_to_border_num < 160) |>
      slice_max(population, n = 10),
    aes(geometry = geometry, label = city),
    stat = "sf_coordinates",
    size = 3)
```

4.3 Most Populous City per State in the Zone

```{r}
label_data <- zone_cities |>
  filter(dist_to_border_num < 160) |>
  group_by(state_name) |>
  slice_max(order_by = population, n = 1, with_ties = FALSE) |>
  ungroup()

ggplot() +
  geom_sf(data = conus_boundary, fill = NA, color = "black", linewidth = 0.4) +
  
  geom_sf(data = zone_cities, aes(color = dist_to_border_num), size = 1) +
  
  scale_color_gradient(low = "orange", high = "darkred") +
  
  gghighlight(dist_to_border_num < 160) +
  
  geom_label_repel(
    data = label_data,
    aes(geometry = geometry, label = city),
    stat = "sf_coordinates",
    size = 3,
    max.overlaps = Inf)
```

In 2–3 sentences: (1) how many states have at least one city in the zone? (2) Are there states where the most populous city is NOT in the zone — what does that tell you about that state’s geography? (3) How might the jurisdictional boundary created by this rule create challenges for environmental monitoring programs that cross state or national boundaries?

According to the map, 23 states have at least one city in the zone. Yes, there are some states where the most populous city may not be within the zone due to availability of resources within the state. This could be an issue as many agencies must come together to monitor the boundaries across states, introducing issue of miscommunication and misalignment through the use of different units and technology.

## Question 5: Interactive Leaflet

```{r}
library(leaflet)

#WGS84 coordinates
us_cities_trans <- st_transform(us_cities, 4326)
conus_trans <- st_transform(conus_boundary, 4326)

pal_border <- colorNumeric(
  palette = "viridis",
  domain = us_cities_trans$dist_to_border)

zone_data <- us_cities_trans |>
  filter(dist_to_border_num < 160)

pal_zone <- colorNumeric(
  palette = "viridis",
  domain = zone_data$dist_to_border)

equi_data <- us_cities_trans |>
  mutate(difference_mexico_canada = abs(dist_to_mexico - dist_to_canada)) |>
  filter(difference_mexico_canada < 100)

pal_equi <- colorNumeric(
  palette = "viridis",
  domain = equi_data$difference_mexico_canada)

#LEAFLET MAP
leaflet() |>
  
  addProviderTiles(providers$CartoDB.Positron) |>
  
  addPolygons(
    data = conus_trans,
    color = "black",
    weight = 1,
    fill = FALSE,
    group = "CONUS") |>
  
  addCircleMarkers( #all cities
    data = us_cities_trans,
    radius = 3,
    stroke = FALSE,
    fillOpacity = 0.7,
    color = ~pal_border(dist_to_border),
    popup = ~paste0(
      "<b>", city_ascii, "</b><br>",
      state_name, "<br>",
      "Population: ", population, "<br>",
      "Dist to border: ", round(dist_to_border, 1)),
    group = "All Cities") |>
  
    addLegend(
    position = "bottomleft",
    pal = pal_border,
    values = us_cities_trans$dist_to_border,
    title = "Distance to US Border",
    group = "All Cities") |>
  
  addCircleMarkers( #100 miles zone
    data = zone_data,
    radius = 3,
    stroke = FALSE,
    fillOpacity = 0.85,
    color = ~pal_zone(dist_to_border),
    popup = ~paste0(
      "<b>", city_ascii, "</b><br>",
      "Population: ", population, "<br>",
      "Dist to border: ", round(dist_to_border, 1)),
    group = "100-mile Zone") |>
  
  addLegend(
    position = "bottomleft",
    pal = pal_zone,
    values = zone_data$dist_to_border,
    title = "100-mile Zone Distance",
    group = "100-mile Zone") |>
  
 #Equidistant zone
  addCircleMarkers(
    data = equi_data,
    radius = 3,
    stroke = FALSE,
    fillOpacity = 0.85,
    color = ~pal_equi(difference_mexico_canada),
    popup = ~paste0(
      "<b>", city_ascii, "</b><br>",
      "Population: ", population, "<br>",
      "Equidistance diff: ", round(difference_mexico_canada, 1)),
    group = "Equidistant Zone") |>

  addLegend(
    position = "bottomleft",
    pal = pal_equi,
    values = equi_data$difference_mexico_canada,
    title = "Equidistance Difference",
    group = "Equidistant Zone") |>
  
  addLayersControl(
    overlayGroups = c(
      "All Cities",
      "100-mile Zone",
      "Equidistant Zone",
      "State Borders"),
    options = layersControlOptions(collapsed = FALSE))
```

In 2–3 sentences, describe the CRS transformation needed to convert from `eqdc` to `leaflet`-compatible coordinates, and any design choices you made to keep the map responsive (e.g., filtering, clustering, or simplifying geometries).

The data was converted from an equal-area projection (eqdu) to WGS84 because Leaflet requires latitude/longitude coordinates. Also, all color gradients were made separately at the beginning to make sure each layer of the map had its own gradient (instead of one gradient shared across maps). A few data frames were made specifically for the map and each section got its own legend/scale in the bottom left corner (note: turn on one layer at a time to see specific legend for each).
